An assessment of the biodiversity - ecosystem function relationship in southern African woodlands

### 2020-01-17 manuscript/manuscript.tex
Going through Kyle's comments

\usepackage{mynotes}

@@

\title{An assessment of the biodiversity - ecosystem function relationship in southern African woodlands}
\author{John L. Godlee}
\date{}

@@

\maketitle
\tableofcontents
\newpage{}

\begin{abstract}
	Studies in natural and experimental settings have shown a positive relationship between tree species diversity and ecosystem functionality but many ecosystems are understudied in this regard. In this study we conducted the first regional estimation of the relationship between tree species diversity and above-ground biomass in southern African woodlands, using a network of \nplots{} woodland plots. We used structural equation modelling (SEM) to determine the relationship between tree species diversity and aboveground biomass, with comparison to the effects of resource availability and along a gradient of stem density. A positive effect of tree species diversity on biomass was demonstrated, with most of this as an indirect effect via woodland structural diversity. We found that the effect of tree species diversity on biomass increases with stem density. Finally, we found that resource availability determines biomass in southern African woodlands largely indirectly via its effect on species diversity.
\end{abstract}

\section{Introduction}

A large number of studies have shown relationships between biodiversity and ecosystem functionality (e.g. \citealt{Liang2016, Cardinale2009}). The strength and direction of these observed Biodiversity - Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{Liang2016}. Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}. %These studies support early findings from small scale experimental studies which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.

Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the degree of biotic activity within an ecosystem \citep{Jax2005}. This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass. The frequently reported and intuitive relationship between biodiversity and ecosystem functionality invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater diversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater diversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater diversity are more likely to include combinations of species which together increase the others functional contribution.

The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages \citep{Barthlott2005}. In wet tropical forests, much of the observed effect of biodiversity on ecosystem function can be attributed to niche complementarity, which is driven by interspecific competition \citep{Wright2017, Poorter2015, Sande2017a}. In the dry tropics however, low precipitation, high seasonality of rainfall and temperature, and high levels of disturbance from fire and herbivory mean that niche complementarity and competition may not play as great a role in the relationship between biodiversity and ecosystem function. Disturbance may weaken the role of competition in determining local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979,Grace1990}. Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{Lasky2014, Tobner2016}. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth \citep{Ratcliffe2017}.

Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016} (Figure \autoref{clust_map}). The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}. Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species. Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}. Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely un-quantified consequences for ecosystem structure and functionality.

Southern African woodlands are relied upon heavily for their ecosystem service provision, which is itself affected by ecosystem functionality \citep{Schulze1994}. This has raised interest in how biodiversity influences ecosystem function in these ecosystems. Resource extraction by humans directly influences biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Aleman2016, Ryan2016}. Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality which heavily influences the degree of seasonal drought and thus woodland structure \citep{Scholes2004, Eldridge2012}. While rapid biodiversity change is being observed in southern African woodlands \citep{Syampungani2009}, research into the relationship between biodiversity and ecosystem functionality remains scarce.

A small number of studies in southern African woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}, but the scientific impact of these studies has been hampered by a restricted climatic and biogeographical range of study sites. Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and woody biomass, it is important to sample across these gradients to be able to infer a truly regional scale relationship between diversity and biomass. Studies conducted over small environmental gradients often find that at local scales, diversity shows a strong effect on ecosystem function, but at large scales diversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}. Small scale studies may therefore be over-estimating the effect of diversity.

In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{Spasojevic2014}. It is important therefore to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on woody biomass in analyses at large spatial scales. \citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation (MAP) sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}. Similarly, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in a wet Panamanian tropical forest. In European forests \citep{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments. They suggest that in dry ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function. Stress tolerance rather than competitiveness may be the mechanism by which a tree species' success is measured in dry forested ecosystems.

@@

\citet{Solbrig1996} writes that southern African woodlands possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy at different growth stages and among species. This structural diversity may be one mechanism through which diversity influences woody biomass. \citep{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest. They also found that trees growing in species rich neighbourhoods exhibited enhanced biomass production. Occupation of multiple canopy layers allows a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area as a form of niche complementarity. This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas, which are structured by disturbance rather than competition. Instead, disturbance history may override the effects of species diversity on structural diversity nullifying the effects of species diversity on structural diversity.

%\citet{Zhang2012} conducted a meta-analysis of 54 studies of the BEFR in forest ecosystems and found that greater tree species abundance evenness led to an increased effect size of species richness on productivity. They suggest this effect is caused by increased heterogeneity of functional traits over space at small scales, leading to a more evenly filled environmental niche volume, which allows a greater number of species to maximise their productivity. We expect that this effect will occur in southern African woodlands as well. A decrease in species abundance evenness may be found as a result of increased disturbance, which prevents dominance by one species, or due to more extreme environmental conditions which preclude the growth of many species, instead favouring those few species which can effectively tolerate sub-optimal conditions. It is important therefore to consider not only species richness as a measure of diversity, but also the evenness of the distribution of trees among species when assessing the impact of diversity on ecosystem function.

%High levels of disturbance in the form of seasonal fires and herbivory from large mammals may moderate the strength of observable BEFRs in southern African woodlands \citep{Staver2009, Bond2005}. Fire disturbance in forests has been linked to abundance dependent mortality among smaller stems \citep{Roques2001}. A dominant and competitive species which tends to produce many small fast-growing stems is more likely to experience mortality during a fire as the higher surface area to volume ratio of the trunk increases the likelihood of combustion. These fast-growing bushy species exist in southern African woodlands and often proliferate when fire is deliberately excluded, leading to the growth of thicket vegetation which precludes the growth of high biomass tree species \citep{Higgins2007}. Additionally, the dominance of a single species is likely to decrease overall functioning by decreasing niche space utilisation \citep{Cardinale2002}.

In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship in southern African woodlands, using inventory plots which span multi-dimensional environmental and biogeographical gradients (\autoref{clust_map}). We investigated the relationship between aboveground woody biomass and tree biodiversity. We compared the relative effects of tree species biodiversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation: precipitation, temperature and soil fertility. We also investigated the potential moderation effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type based on major tree species compositional units as a factor in our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality. Initially, we made three hypotheses:

\subsection{Hypotheses}

\begin{enumerate}
	\item{Precipitation and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity.}
	\item{The effect size of species diversity on woody biomass will increase with stem density, due to an increased importance of niche complementarity as competition increases.}
	\item{Tree structural diversity will indirectly influence tree species diversity to provide an indirect path of influence between structural diversity and woody biomass.} \label{lst:num}
\end{enumerate}

%To address these hypotheses we used Structural Equation Models with path analysis as a preferred method to simultaneously account for environmental factors and biotic factors, which may interact in their effect on ecosystem structure and therefore biomass.

\section{Methods}

\subsection{Study location}

The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in so-called miombo woodlands (\autoref{clust_map}, \citealt{White1987}). The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. The study sites span the core climate space of the region as a whole. The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland area as defined by \citet{White1987}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.

@@

\subsection{Data collection}

@@

\subsection{Data analysis}
Estimated tree species richness was calculated for each plot using \verb|ChaoRichness()| from the \verb|iNEXT| package in R \citep{Hsieh2016}. This procedure extrapolates a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value. Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort among plots. Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}.

@@

\subsubsection{Vegetation clusters}

@@

\begin{figure}[H]
\centering
	\includegraphics[width=1.4\textwidth]{clust_map}
	\caption{The locations of the \nplots{} plots used in this study, as points by geographic location with respect to the distribution of miombo woodland vegetation according to \citet{White1987}. Each panel shows plots categorized by their vegetation type as defined by the vegetation types in \autoref{clust_summ}.}
	\label{clust_map}
\end{figure}
\end{landscape}

\subsubsection{Structural Equation Modelling}

@@

\section{Results}

@@

\subsection{Structural and species diversity models}

@@

\subsection{Variation among vegetation types}

@@

\subsection{Moderation of Diversity-AGB relationship by stem density}

@@

\subsection{Environmental covariates and diversity}

@@

% \subsection{Moderation effects of moisture availability and soil}
% 
% To address the hypothesis (H\textsubscript{1}) that more arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above ground woody biomass, we fit separate linear multiple regressions with the latent variables of moisture availability and soil fertility as moderators on the relationship between species diversity and AGB.
% 
% Both soil fertility and moisture availability had small positive interaction effects on the strength of the relationship between tree species diversity and biomass (\autoref{int_plots}). The regression coefficient for the interaction effect of tree species diversity and moisture availability was significant (\moderp{}), while the interaction term of tree species diversity and soil fertility was marginally significant (\moders{}). As moisture availability increased, the effect of tree species diversity on AGB became stronger. See \autoref{mois_div_int_mod} and \autoref{soil_div_int_mod} for full regression tables.
% 
% 
% \begin{figure}[H]
% \centering
% 	\subfloat[]{
% 	\includegraphics[width=0.48\textwidth]{mois_int}
% 	}
% 	\subfloat[]{
% 	\includegraphics[width=0.48\textwidth]{soil_int}
% 	}
% 	\caption{Scatter plots showing variation in the relationship between tree species diversity and AGB with moisture availability (a) and soil fertility (b). All variables are standardised and centred to a mean of zero and a standard deviation of 1. Lines of best fit are drawn for standard deviations of moisture availability and soil fertility between -2 and +2. Moisture availability, tree species diversity and soil fertility are latent variables comprised of the same observed variable loadings.}
% 	\label{int_plots}
% \end{figure}

\section{Discussion}

@@

In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] resource availability via moisture availability and soil fertility, [d] stem density and their interactions on above ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland survey plots. Using latent variables and Structural Equation Modelling (SEM), we found support for a general relationship between tree species diversity and AGB, with an indirect relationship between tree species diversity and AGB via structural diversity (H\textsubscript{1}). We found that the effect size of tree species diversity on AGB increased with stem density (H\textsubscript{2}).  Tree diversity, structural diversity and stem density accounted for \srsq{}\% of the variation in AGB across the region, while models for certain vegetation types showed even greater explanatory power (\autoref{struc_model_fit_clust_stats}). The strongest effect on AGB was that of stem density. Interestingly, when tree species diversity, structural diversity and stem density were controlled for, we found little evidence of an effect of resource availability, in the form of moisture availability and soil fertility, on AGB (H\textsubscript{3}). We found that vegetation composition in the form of discrete vegetation types affected the relationship between tree species diversity and AGB.

\subsection{Effect of tree species diversity on AGB}

% I would just write that you had a consistent positive effect across all models, as that is your general result with regard to this question.

We found a consistent positive total effect of tree species diversity on AGB across all models in this study. Within the savanna woodlands of southern Africa we therefore find support that higher tree species richness causes higher woody AGB. This finding is in agreement with many other studies across different ecosystems and biomes, showing that there is a generalisable positive association between species richness and ecosystem functionality \citep{Liang2016, Cardinale2009}. In addition however, our study provides a novel dissection of the mechanisms underlying this relationship and their interaction with resource availability and vegetation type in southern African woodlands as an understudied biome.

>>>>

% Define page geometry
\usepackage{geometry}
\geometry{left=2.2cm,
	right=2.2cm,
	top=2.2cm,
	bottom=2cm}
\parskip 0.15cm
\setlength{\parindent}{0cm}
\usepackage{pdflscape}

% Set font
\usepackage[T1]{fontenc}

% Image handling
\usepackage{graphics}  % Insert images easily
\usepackage{graphicx}  % Extended image support

\makeatletter
	\g@addto@macro\@floatboxreset\centering  % Automatically centre images (floats)
\makeatother

\graphicspath{ {img/} }
\usepackage{float}  %  Graphics placement [H] [H!] arguments
\usepackage{subfig}  % Compound figures

% Tables
\usepackage{multirow}
\usepackage{longtable}

% Bibliography management
\usepackage{natbib}    % Bibliography management - Use author/date citations
\bibliographystyle{agsmnourl}  % Use custom agsm bibliography template with no URL
\usepackage{cite}  % Citation options

% Text formatting
\usepackage{url} % Allow nice formatting of URLs in text

\usepackage{enumerate}  % Enumerated lists

\usepackage{lineno}  % Line numbers

\usepackage{textcomp}
\newcommand{\textapprox}{\raisebox{0.5ex}{\texttildelow}}  % Command for a good tilde

\usepackage{siunitx}
\usepackage{amsmath}

\usepackage{xcolor}
\newcommand{\todo}[1]{\textcolor{red}{\textbf{#1}}}   % \todo{NOTE TO SELF WRITTEN IN RED}

\input{code_format}

% Custom title formatting
\let\oldtitle\title

\renewcommand{\title}[1]{\oldtitle{\vspace{-1.5cm}#1}}

\usepackage[breaklinks]{hyperref}
\definecolor{links}{RGB}{191,59,72}
\hypersetup{
	breaklinks,
	colorlinks,
	allcolors=links,
	linktoc=section,
	pdfauthor={John L. Godlee}
}
\def\subsectionautorefname{section}
\def\subsubsectionautorefname{section}
% \usepackage{biblatex}

@@

\usepackage{lineno}
\linenumbers

\title{An assessment of the biodiversity - ecosystem function relationship in southern African woodlands}
\author{}
\date{}

@@

{\LARGE{Title: An assessment of the biodiversity - ecosystem function relationship in southern African woodlands}}

\vspace{1cm}

Authors: Godlee, J. L.\textsuperscript{1}, Dexter, K. G.\textsuperscript{1}, Ryan, C. M.\textsuperscript{1}

\textsuperscript{1}: School of GeoSciences, University of Edinburgh, Edinburgh, United Kingdom \\
\textsuperscript{2}: Some other address

\vspace{1em}
Corresponding author:

John L. Godlee

johngodlee@gmail.com

School of GeoSciences, University of Edinburgh, Edinburgh, United Kingdom

\section*{Acknowledgements}

\section*{Biosketch}
\newpage{}

{\LARGE{\textbf{Blinded Main Text File}}}

\LARGE{Title: An assessment of the biodiversity - ecosystem function relationship in southern African woodlands}

\normalsize{Running title: Woodland functionality in southern Africa}

\section*{Abstract}

\textbf{Aim:} Positive correlations between tree biodiversity and ecosystem function have been widely documented, but the nature of the relationship in highly disturbed and ecophysiologically stressful systems is less clear. Here, we explore the relationship between tree biodiversity and aboveground biomass across southern African savannas and woodlands, with respect to gradients in stem density and resource availability, to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.

\textbf{Location:} Southern African savannas and woodlands

\textbf{Time period:} 2010-2019

\textbf{Major taxa studied:} Trees

\textbf{Methods:} We used a network of \nplots{} savanna and woodland plots located across the southern African sub-continent in which each tree >5 cm diameter was measured. We used Structural Equation Modelling and path analysis to determine the relationship between tree species diversity and aboveground biomass, with comparison to the effects of resource availability and along a gradient of stem density.

\textbf{Results:} A positive effect of tree species diversity on biomass was demonstrated, observed largely as an indirect effect of increasing woodland structural diversity. We also found that the effect of tree species diversity on biomass increases with stem density. Finally, we found that resource availability affects biomass in southern African woodlands largely indirectly via its effect on species diversity.

\textbf{Main conclusions:} Tree species diversity plays an important role in determining the ecosystem functionality and structure of southern African woodlands. Our results demonstrate the importance of considering environmental conditions and woodland structural attributes to more accurately describe the relationship between biodiversity and ecosystem functionality.

% Studies in natural and experimental settings have shown a positive relationship between tree species diversity and ecosystem functionality but many ecosystems are understudied in this regard. In this study we conducted the first regional estimation of the relationship between tree species diversity and above-ground biomass in southern African woodlands, using a network of \nplots{} woodland plots. We used structural equation modelling (SEM) to determine the relationship between tree species diversity and aboveground biomass, with comparison to the effects of resource availability and along a gradient of stem density. A positive effect of tree species diversity on biomass was demonstrated, with most of this as an indirect effect via woodland structural diversity. We found that the effect of tree species diversity on biomass increases with stem density. Finally, we found that resource availability determines biomass in southern African woodlands largely indirectly via its effect on species diversity.

\section*{Introduction}

Numerous studies have shown relationships between biodiversity and ecosystem function (e.g. \citealt{Liang2016, Cardinale2009}). The strength and direction of these observed Biodiversity - Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{Liang2016}. Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}. These studies support early findings from small scale experimental studies predominantly in grassland patches, which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.

Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the degree of biotic activity within an ecosystem \citep{Jax2005}. This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass. The frequently reported and intuitive relationship between biodiversity and ecosystem functionality invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater diversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater diversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater diversity are more likely to include combinations of species which together increase the others' functional contribution.

Compared to other forested ecosystems, dry tropical woodlands are highly structured by disturbance, mainly through fire, and in Africa notably by herbivory also \citep{}. High levels of disturbance may weaken the role of competition in determinging local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}. This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as apparent in dry woodland ecosystems. Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{Lasky2014, Tobner2016}. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth. In European forests \citep{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments. They suggest that in dry ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function. This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation.

% COME BACK TO THIS: In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{Spasojevic2014}. It is important therefore to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on woody biomass in analyses at large spatial scales. \citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation (MAP) sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}. Similarly, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in a wet Panamanian tropical forest.

The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages \citep{Barthlott2005}. A severe lack of study in dry tropical ecosystems, especially given the potential mismatch in BEFR mechanism described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEFR and it's environmental determinants.

Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016} (\autoref{clust_map}). The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}. Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species. Southern African savannas and woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}. Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely un-quantified consequences for ecosystem structure and functionality.

A small number of studies in southern African savannas and woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. However, studied conducted over small environmental gradients often find that at local scales, diversity shows a strong effect on ecosystem function, but at large scales diversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}. Small scale studies may therefore be over-estimating the effect of diversity. Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and woody biomass \citep{}, it is important to sample across geographic and environmental gradients to be able to infer a truly regional scale relationship between diversity and biomass.

@@

\citet{Solbrig1996} writes that southern African woodlands possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy at different growth stages and among species. This structural diversity may be one mechanism through which diversity influences woody biomass. \citep{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest. They also found that trees growing in species rich neighbourhoods exhibited enhanced biomass production. Occupation of multiple canopy layers allows a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area via a form of niche complementarity. This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas, which are structured by disturbance rather than competition. Instead, disturbance history may override the effects of species diversity on structural diversity nullifying the effects of species diversity on structural diversity.

% High levels of disturbance in southern African woodlands may moderate the observable BEFR through it's effect on ecosystem composition. Fire disturbance in forests has been linked to abundance dependent mortality among smaller tree stems \citep{Roques2001, Staver2009, Bond2005}. A dominant and competitive species which tends to produce many small fast-growing stems is more likely to experience mortality during a fire as the higher surface area to volume ratio of the trunk increases the likelihood of combustion. These fast-growing bushy species often proliferate in southern African woodlands when fire is deliberately excluded, leading to the growth of thicket vegetation which precludes the growth of high biomass tree species \citep{Higgins2007}. Additionally, the dominance of a single species is likely to decrease overall functioning by decreasing niche space utilisation \citep{Cardinale2002}.
% KGD: This text is useful in demonstrating your biological understanding of the ecosystem. I think it could be useful to include some of this biological knowledge in the paragraph above that ends with the statements on disturbance (the paragraph starting with Slobrig 1996). Or you could include this whole paragraph just after that one. The co-authors will reign us in if this kind of logic is not sensible for the ecosystems we are studying.

In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship across southern African savannas and woodlands, using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}). We compared the relative effects of tree species biodiversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility. We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type based on major tree species compositional units as a factor in our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality and assess the generality of our results. We used Structural Equation Modelling (SEM) and path analysis as a preferred method to simultaneously account for environmental and biotic factors, which may interact their effect on ecosystem structure and therefore biomass. Initially, we made three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect size of species diversity on woody biomass will increase with stem density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity, which will provide an indirect path by which tree diversity increases woody biomass.

\section*{Materials and methods}

\subsection*{Study location}

The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in so-called miombo woodlands (\autoref{clust_map}, \citealt{White1983}). The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. The study sites span the core climate space of the region as a whole. The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland area as defined by \citet{White1983}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.

@@

\subsection*{Data collection}

@@

\subsection*{Data analysis}
Estimated tree species richness was calculated for each plot using \verb|ChaoRichness()| from the \verb|iNEXT| package in R \citep{Hsieh2016}. This procedure extrapolates a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value. Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort among plots. Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}.

@@

\subsubsection*{Vegetation clusters}

@@

\begin{figure}[H]
\centering
	\includegraphics[width=1.4\textwidth]{clust_map}
	\caption{The locations of the \nplots{} plots used in this study, as points by geographic location with respect to the distribution of miombo woodland vegetation according to \citet{White1983}. Each panel shows plots categorized by their vegetation type as defined by the vegetation types in \autoref{clust_summ}.}
	\label{clust_map}
\end{figure}
\end{landscape}

\subsubsection*{Structural Equation Modelling}

@@

\section*{Results}

@@

\subsection*{Structural and species diversity models}

@@

\subsection*{Variation among vegetation types}

@@

\subsection*{Moderation of Diversity-AGB relationship by stem density}

@@

\subsection*{Environmental covariates and diversity}

@@

% \subsection*{Moderation effects of moisture availability and soil}
% 
% To address the hypothesis (H\textsubscript{1}) that more arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above ground woody biomass, we fit separate linear multiple regressions with the latent variables of moisture availability and soil fertility as moderators on the relationship between species diversity and AGB.
% 
% Both soil fertility and moisture availability had small positive interaction effects on the strength of the relationship between tree species diversity and biomass (\autoref{int_plots}). The regression coefficient for the interaction effect of tree species diversity and moisture availability was significant (\moderp{}), while the interaction term of tree species diversity and soil fertility was marginally significant (\moders{}). As moisture availability increased, the effect of tree species diversity on AGB became stronger. See \autoref{mois_div_int_mod} and \autoref{soil_div_int_mod} for full regression tables.
% 
% 
% \begin{figure}[H]
% \centering
% 	\subfloat[]{
% 	\includegraphics[width=0.48\textwidth]{mois_int}
% 	}
% 	\subfloat[]{
% 	\includegraphics[width=0.48\textwidth]{soil_int}
% 	}
% 	\caption{Scatter plots showing variation in the relationship between tree species diversity and AGB with moisture availability (a) and soil fertility (b). All variables are standardised and centred to a mean of zero and a standard deviation of 1. Lines of best fit are drawn for standard deviations of moisture availability and soil fertility between -2 and +2. Moisture availability, tree species diversity and soil fertility are latent variables comprised of the same observed variable loadings.}
% 	\label{int_plots}
% \end{figure}

\section*{Discussion}

@@

In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] resource availability, i.e. moisture availability and soil fertility, [d] stem density and their interactions on above ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland survey plots. Using latent variables and Structural Equation Modelling (SEM), we found support for a general positive relationship between tree species diversity and AGB, with an indirect influence of tree species diversity on AGB via structural diversity (H\textsubscript{1}). We found that the effect size of tree species diversity on AGB increased with stem density (H\textsubscript{2}).  Tree diversity, structural diversity and stem density accounted for \srsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power (\autoref{struc_model_fit_clust_stats}). The strongest effect on AGB was that of stem density. Interestingly, when the effects of tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, in the form of moisture or energy availability or soil fertility, on AGB (H\textsubscript{3}).

\subsection*{Inter-related effects of tree species and structural diversity on AGB}

We found a consistent positive effect of tree species diversity on AGB across all models in this study. Within the savanna woodlands of southern Africa we therefore find support that higher tree species richness and evenness causes higher woody AGB. This finding is in agreement with many other studies across different ecosystems and biomes, supporting that there is a generalisable positive association between diversity and ecosystem functionality \citep{Liang2016, Cardinale2009}. Our study provides a novel dissection of the mechanisms underlying this relationship, particularly in the context of a disturbance-structured  and poorly studied ecological system, southern African woodlands and savannas.

% In our extrapolation of species richness according to sampling effort (stem density) we partially accounted for this effect, but our analyses cannot properly assess to what extent a real effect of species diversity on stem density exists. We cannot decompose the relative effects of tree species richness and abundance evenness in our model, but previous studies have shown that both richness and evenness have similar importance in their effects on ecosystem function \citep{Valery2009, Zhang2012}.
